---
title: "Practical Machine Learning Course Project"
author: "Anna Teplukhina | `r Sys.Date()`"
#date: "`r Sys.Date()`"
output: html_document
---

## Overview
This project is dedicated to analysis

## Pre-modelling setup and data preparation 
First one can load all libraries that will be required for the following analysis and to set a seed to ensure reproducibility:
```{r, echo=TRUE,warning=FALSE}
library(lattice)
library(ggplot2)
library(caret)

set.seed(135)
```

There are two available data sets: training one and testing one. One can load the data and check the data structure:
```{r}
training <- read.csv("pml-training.csv")
testing <- read.csv("pml-testing.csv")

dim(training)
dim(testing)
```

In total, there are 160 variables and 19622/20 observations in the training/testing sets.

Before starting to develop a predictive model, it is necessary to clean the training data set and to remove unessential features, such as variables with almost zero variance and NA variables. Also, first 5 variables (```X```, ```user_name```, ```raw_timestamp_part_1```, ```raw_timestamp_part_2```, ```cvtd_timestamp```) can be excluded too since they do not contribute much to analysis.
```{r}
# Removing variables with almost zero variance
tr_nzv <- nearZeroVar(training)
training_cl <- training[,-tr_nzv]

# Removing NA variables
training_cl <- training_cl[ ,colMeans(is.na(training_cl)) < .9]

# Removing first 5 variables
training_cl <- training_cl[,-c(1:5)]

dim(training_cl)
```
